The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-10 rational.tex
conclusion

@@

To summarize negative, fractional, square root, and imaginary types all make
sense. What they help you accomplish as a programmer is to disassociate
global invariants into local ones that can be satisfied independently by
subcomputations with no synchronization or communication. A computation
producing something of type $a/b$ does not need to concern itself with who is
going to supply the missing $b$: it just does its part. Conversely faced with
a complicated task, a computation might decide to break it into pieces and
demand these pieces using negative types.

It is no surprise that these types are closely related to quantum mechanics
and that they give us the feel that this is how nature computes. This is
speculation however.

In any case, in a framework where information can be copied and deleted, none
of this makes much sense. It is critical that these constraints and demands
can neither be duplicated nor erased.  This gives us the maximum
``parallelism'' possible.

Say we have not considered recursion in this paper.

The simplest way to connect the In a conventional computational model, one
might realize this situation by simply writing the identity function: the
buyer hands the money to the seller to finish the transaction. The above
sequence of isomorphisms implements this identity function in a much more
interesting way, however.  of the above series of

On the producer side, the debt is paid for by the money computation with the
identity function. the producer and consumer must somehow share an explicit
dependency that allows the value. However in our model, the presence of
negative types allows the produced value to satisfy the demand without the
producer or consumer even knowing about each other. As is explained in detail
in

Furthermore, we illustrate their true appeal and expressiveness is brought
forth by viewing them in the context of an information-preserving
computational model.

In addition, we show how these types enrich our computational model, they
obey the same laws as the rational numbers. In other words, every algebraic
identity that holds for the rational numbers corresponds to a type
isomorphism in our model.

Specifically, isomorphisms enrich our computational model with have an
interesting computational interpretation

@@

We furthermore demonstrate that, in our model,
programming with these negative and fractional types, is a new ``revolution''
breaking dependencies...

@@

Computationally, symmetry exhibits itself as a duality between two concepts.

Symmetry is pervasive in both natural and man-made environments.

@@

We introduced this thesis that computation should be based on isomorphisms
that preserve information~\cite{infeffects}. Since Filinski, we've had the
idea that values and continuations are like mirror images. In a conventional
language, the negative (continuation) side is implicit and we introduce
information effects on the positive. Trying to recover the duality from this
distorted positive side has always been messy. Now it looks clean because we
have kept the positive side pure.

In a technical sense, this paper extends the language of isomorphisms
{{langRev}}, with duality. Unlike Linear logic \cite{Girard87tcs} and
other systems which have one notion of duality over additive and
multiplicative components, {{langRevEE}} has two notions of duality
-- an additive duality over the monoid {{(0, +)}} and a multiplicative
duality over the monoid {{(1, *)}}. Each axis of duality also give us
a function space and hence {{langRevEE}} has an additive function
space corresponding to a notion of control or backtracking and a
multiplicative function space corresponding to a notion of unification
or constraint satisfaction.

The world of computation we are describing has:
\begin{itemize}
\item suppliers, 
\item consumers, and
\item bi-directional transformations
\end{itemize}
This is the same world described by the papers on the duality of computation
but that work only scratched the surface! We have the following features:
\begin{itemize}
\item we can start from the supplier and push the values towards the
  consumer (call-by-value in the duality of computation papers)
\item we can start from the consumer and pull the values from the suppliers
  (call-by-name in the duality of computation papers)
\item we can combine the pushing and pulling and values using eta/epsilon for
  sum types; these allow us to at any point in the middle of the computation
  create out of nothing a value to send to the consumers and a demand to send
  to the suppliers.
\item we can break a big data structure into fragments described by
  fractional types; the suppliers and consumers can produce and consume the
  pieces completely independently of each other. Eventually the pieces will
  fit together at the consumer to produce the desired output.
\item we can break a bi-directional transformation into pieces using square
  roots
\item we can take into account that values have phase (complex numbers),
  i.e., it is not that they flow towards the consumer or just towards the
  suppliers; they can be flowing in direction that ``30 degrees'' towards the
  consumer for example.
\end{itemize}
So it is all about breaking dependencies in some sense to allow for maximum
autonomy (parallelism) of subcomputations. It is probably the case that to
make full use of square root types and imaginary types, we have to move to a
vector space. If that's the case, we should probably leave this stuff out and
focus on negative and fractional types and only have a short discussion of
the polynomials restricted to seven trees in one and similar issues.

The conventional idea is to divide the world into a ``real'' one and a
``virtual'' one. In the ``real'' world, we can define datatypes like
\verb|t=t^2+1| but we don't have additive inverses so it makes no sense to
talk of negative types and we can't rearrange the terms in the datatype
definition. However the observation is that we can map these datatypes to a
virtual world that has more structure (a ring that provides additive inverses
or a field that also provides multiplicative inverses) and then perform
computations in the ring/field. If we perform computations in the ring, then
some of these will use additive inverses in ways that cannot be mapped back
to the ``real'' world. Much of current research attempts to characterize
which computations done in the ring are valid isomorphisms between datatypes
in the ``real'' world. This is nice but is not what I am after. In fact I am
not interested in the ring or the semiring at all. I am interested in the
field and I want this field to be \textbf{the real world.} This is partly
motivated by the fact that Quantum Mechanics seems to demand an underlying
field and more generally that the field provides the maximum generality in
slicing and dicing computations. So assuming I live in a field and that the
negative, fractional, square root, and imaginary types are all ``real,'' how
do I compute in this field? Clearly there will be constraints on
``measurement'' in the sense that a full program cannot produce any of the
crazy types but that's done outside the formalism in some sense just as in
Quantum Computing. The main question I am after is how to compute in this
field with first-class negative, fractional, etc. types. As I mentioned in my
previous email, we can produce programs that have types \verb|t^3 <-> -1| and
they ``run'' (but only to give infinite loops).

So when a programmer writes the datatype declaration \verb|t = t^2+1|,
if we allow negative etc. then this is effectively writing
\verb|t = cubicroot{-1}|. If we are in the field then computations
that manipulate these trees can be sliced and diced even at interfaces
that expose the cubic root and the imaginary types.

Future work: develop a type system for a ``normal language'' that has
negative, fractional, etc. types as first-class types. More long term,
instead of adding one polynomial at a time, we can go to an algebraically
closed field. The complex numbers is an obvious choice but I would rather go
to something computable like the field of algebraic numbers. Is the adele
ring or the p-adics relevant here?

@@

We introduced this thesis that computation should be based on isomorphisms
that preserve information~\cite{infeffects}. Since Filinski, we've had the
idea that values and continuations are like mirror images. In a conventional
language, the negative (continuation) side is implicit and we introduce
information effects on the positive. Trying to recover the duality from this
distorted positive side has always been messy. Now it looks clean because we
have kept the positive side pure.

The way to think about something of type $A$ is that it is a value we have
produced. The way to think about something of type $-A$ is that is a value we
have already consumed.

>>>>

We have extended the language {{langRev}} that expressed computation
in the commutative semiring of whole numbers to {{langRevEE}} that
expresses computation in the fields of rationals.  Every algebraic
identity that holds for the rational numbers corresponds to a type
isomorphism in our model.  In the process, we have examined the two
function spaces that these give rise to an have seen non-trivial
constructions such as a novel SAT-solver that relies on a
multiplicative trace.

In another sense however, this paper is about the nature of duality in
computation. The concept of duality is deep and significant and in
many cases -- probably most famously in the divide between classical
and intuinionistic mathematics -- contain non-trivial assumptions
about our world-view.

The problem of duality in computation and logic, has many facets of
which `continuation' and `higher order functions' are only one.

LL, GoI

Int Construction

QC

Parallel computing with constraints and backtracking -- breaking
dependancies,

We have contributed an elegant and natural notion of duality that
arises from the monoidal structure of sums and pairs. This notion of
duality has a crisp semantics, clear computational interpretations and
an information theoretic basis.

--- END --

@@

%% To summarize negative, fractional, square root, and imaginary types all make
%% sense. What they help you accomplish as a programmer is to disassociate
%% global invariants into local ones that can be satisfied independently by
%% subcomputations with no synchronization or communication. A computation
%% producing something of type $a/b$ does not need to concern itself with who is
%% going to supply the missing $b$: it just does its part. Conversely faced with
%% a complicated task, a computation might decide to break it into pieces and
%% demand these pieces using negative types.

%% It is no surprise that these types are closely related to quantum mechanics
%% and that they give us the feel that this is how nature computes. This is
%% speculation however.

%% In any case, in a framework where information can be copied and deleted, none
%% of this makes much sense. It is critical that these constraints and demands
%% can neither be duplicated nor erased.  This gives us the maximum
%% ``parallelism'' possible.

%% Say we have not considered recursion in this paper.

%% The simplest way to connect the In a conventional computational model, one
%% might realize this situation by simply writing the identity function: the
%% buyer hands the money to the seller to finish the transaction. The above
%% sequence of isomorphisms implements this identity function in a much more
%% interesting way, however.  of the above series of

%% On the producer side, the debt is paid for by the money computation with the
%% identity function. the producer and consumer must somehow share an explicit
%% dependency that allows the value. However in our model, the presence of
%% negative types allows the produced value to satisfy the demand without the
%% producer or consumer even knowing about each other. As is explained in detail
%% in

%% Furthermore, we illustrate their true appeal and expressiveness is brought
%% forth by viewing them in the context of an information-preserving
%% computational model.

%% In addition, we show how these types enrich our computational model, they
%% obey the same laws as the rational numbers.

%% Specifically, isomorphisms enrich our computational model with have an
%% interesting computational interpretation

@@

%% We furthermore demonstrate that, in our model,
%% programming with these negative and fractional types, is a new ``revolution''
%% breaking dependencies...

@@

%% Computationally, symmetry exhibits itself as a duality between two concepts.

%% Symmetry is pervasive in both natural and man-made environments.

@@

%% We introduced this thesis that computation should be based on isomorphisms
%% that preserve information~\cite{infeffects}. Since Filinski, we've had the
%% idea that values and continuations are like mirror images. In a conventional
%% language, the negative (continuation) side is implicit and we introduce
%% information effects on the positive. Trying to recover the duality from this
%% distorted positive side has always been messy. Now it looks clean because we
%% have kept the positive side pure.

%% In a technical sense, this paper extends the language of isomorphisms
%% {{langRev}}, with duality. Unlike Linear logic \cite{Girard87tcs} and
%% other systems which have one notion of duality over additive and
%% multiplicative components, {{langRevEE}} has two notions of duality
%% -- an additive duality over the monoid {{(0, +)}} and a multiplicative
%% duality over the monoid {{(1, *)}}. Each axis of duality also give us
%% a function space and hence {{langRevEE}} has an additive function
%% space corresponding to a notion of control or backtracking and a
%% multiplicative function space corresponding to a notion of unification
%% or constraint satisfaction.

The world of computation we are describing has:
\begin{itemize}
\item suppliers, 
\item consumers, and
\item bi-directional transformations
\end{itemize}
This is the same world described by the papers on the duality of computation
but that work only scratched the surface! We have the following features:
\begin{itemize}
\item we can start from the supplier and push the values towards the
  consumer (call-by-value in the duality of computation papers)
\item we can start from the consumer and pull the values from the suppliers
  (call-by-name in the duality of computation papers)
\item we can combine the pushing and pulling and values using eta/epsilon for
  sum types; these allow us to at any point in the middle of the computation
  create out of nothing a value to send to the consumers and a demand to send
  to the suppliers.
\item we can break a big data structure into fragments described by
  fractional types; the suppliers and consumers can produce and consume the
  pieces completely independently of each other. Eventually the pieces will
  fit together at the consumer to produce the desired output.
\item we can break a bi-directional transformation into pieces using square
  roots
\item we can take into account that values have phase (complex numbers),
  i.e., it is not that they flow towards the consumer or just towards the
  suppliers; they can be flowing in direction that ``30 degrees'' towards the
  consumer for example.
\end{itemize}

%% So it is all about breaking dependencies in some sense to allow for maximum
%% autonomy (parallelism) of subcomputations. It is probably the case that to
%% make full use of square root types and imaginary types, we have to move to a
%% vector space. If that's the case, we should probably leave this stuff out and
%% focus on negative and fractional types and only have a short discussion of
%% the polynomials restricted to seven trees in one and similar issues.

%% The conventional idea is to divide the world into a ``real'' one and a
%% ``virtual'' one. In the ``real'' world, we can define datatypes like
%% \verb|t=t^2+1| but we don't have additive inverses so it makes no sense to
%% talk of negative types and we can't rearrange the terms in the datatype
%% definition. However the observation is that we can map these datatypes to a
%% virtual world that has more structure (a ring that provides additive inverses
%% or a field that also provides multiplicative inverses) and then perform
%% computations in the ring/field. If we perform computations in the ring, then
%% some of these will use additive inverses in ways that cannot be mapped back
%% to the ``real'' world. Much of current research attempts to characterize
%% which computations done in the ring are valid isomorphisms between datatypes
%% in the ``real'' world. This is nice but is not what I am after. In fact I am
%% not interested in the ring or the semiring at all. I am interested in the
%% field and I want this field to be \textbf{the real world.} This is partly
%% motivated by the fact that Quantum Mechanics seems to demand an underlying
%% field and more generally that the field provides the maximum generality in
%% slicing and dicing computations. So assuming I live in a field and that the
%% negative, fractional, square root, and imaginary types are all ``real,'' how
%% do I compute in this field? Clearly there will be constraints on
%% ``measurement'' in the sense that a full program cannot produce any of the
%% crazy types but that's done outside the formalism in some sense just as in
%% Quantum Computing. The main question I am after is how to compute in this
%% field with first-class negative, fractional, etc. types. As I mentioned in my
%% previous email, we can produce programs that have types \verb|t^3 <-> -1| and
%% they ``run'' (but only to give infinite loops).

%% So when a programmer writes the datatype declaration \verb|t = t^2+1|,
%% if we allow negative etc. then this is effectively writing
%% \verb|t = cubicroot{-1}|. If we are in the field then computations
%% that manipulate these trees can be sliced and diced even at interfaces
%% that expose the cubic root and the imaginary types.

%% Future work: develop a type system for a ``normal language'' that has
%% negative, fractional, etc. types as first-class types. More long term,
%% instead of adding one polynomial at a time, we can go to an algebraically
%% closed field. The complex numbers is an obvious choice but I would rather go
%% to something computable like the field of algebraic numbers. Is the adele
%% ring or the p-adics relevant here?

@@

%% We introduced this thesis that computation should be based on isomorphisms
%% that preserve information~\cite{infeffects}. Since Filinski, we've had the
%% idea that values and continuations are like mirror images. In a conventional
%% language, the negative (continuation) side is implicit and we introduce
%% information effects on the positive. Trying to recover the duality from this
%% distorted positive side has always been messy. Now it looks clean because we
%% have kept the positive side pure.

%% The way to think about something of type $A$ is that it is a value we have
%% produced. The way to think about something of type $-A$ is that is a value we
%% have already consumed.
